---
title: "Cornell Gorges"
subtitle: "A Data-Driven Look at the Ithacan Gorges"
author: "Ming DeMers"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

# Introduction

The gorges at Cornell University are one of the university's most iconic and scenic features. Nestled in the heart of the campus, Cornell's gorges are a series of deep and narrow ravines carved by the flow of water over thousands of years. These beautiful natural formations offer a tranquil escape from the hustle and bustle of campus life, providing students and visitors with picturesque spots for relaxation and exploration. While the gorges are captivating, they can be deceptively dangerous. Safety measures are enforced and visitors are urged to follow the guidelines. 

The Gorge Stewards Program was established in 2014 to improve safety and enjoyment of the gorges, following the death of a student earlier that year. Gorge stewards walk the paths from May to September to provide information about trails, safety rules, natural history, activities, and swimming alternatives. Stewards also track visitor use, including unauthorized or illegal uses.

This project will explore how usage of the Gorges has fluctuated, how violations of gorge rules has decreased, and explain overall the changing landscape of the Gorges.

# The Data

The data is sourced from the Gorge Stewards Program, courtesy of the Cornell Botanic Gardens and Cornell Outdoor Education (COE). It comprises of a table for each year recorded, with a number of variables detailing gorge usage, violations, weather and more. 

```{r setup, include=FALSE}
#| label: import-load

library(tidyverse)
library(tidylog)
library(janitor)
# knitr::opts_chunk$set(echo = TRUE)

gorge_2015_raw <- read.csv('data/Gorge Steward Log  - 2015 Season.csv')
gorge_2016_raw <- read.csv('data/Gorge Steward Log  - 2016 Season.csv')
gorge_2017_raw <- read.csv('data/Gorge Steward Log  - 2017 Season.csv')
gorge_2018_raw <- read.csv('data/Gorge Steward Log  - 2018 Seson.csv')
gorge_2019_raw <- read.csv('data/Gorge Steward Log  -  2019 Season.csv')
gorge_2015_raw <- read.csv('data/Gorge Steward Log  - 2015 Season.csv')
gorge_2020_raw <- read.csv('data/Gorge Steward Log  - 2020 seaon.csv')
gorge_2021_raw <- read.csv('data/Gorge Steward Log  - 2021 Season.csv')
gorge_2022_raw <- read.csv('data/Gorge Steward Log  - 2022 Season.csv')
gorge_2023_raw <- read.csv('data/Gorge Steward Log  - 2023 Season.csv')

gorge_2015_raw <- gorge_2015_raw |>
  mutate(observed_gorge_users = Total.number.of.observed.gorge.users) |>
  mutate(observed_violations= Total.number.of.observed.violations)

gorge_2016_raw <- gorge_2016_raw |>
  mutate(observed_gorge_users = Total.number.of.observed.gorge.users) |>
  mutate(observed_violations= Total.number.of.observed.violations)

gorge_2017_raw <- gorge_2017_raw |>
  mutate(observed_gorge_users = Total.number.of.observed.gorge.users) |>
  mutate(observed_violations= Total.number.of.observed.violations)

gorge_2018_raw <- gorge_2018_raw |>
  mutate(observed_gorge_users = Total.number.of.observed.gorge.users) |>
  mutate(observed_violations= Total.number.of.observed.violations)

gorge_2019_raw <- gorge_2019_raw |>
  mutate(observed_gorge_users = Total.number.of.observed.gorge.users) |>
  mutate(observed_violations= Total.number.of.observed.violations)

gorge_2020_raw <- gorge_2020_raw |>
  mutate(observed_gorge_users = Total.number.of.observed.gorge.users..Daily.) |>
  mutate(observed_violations = Total.number.of.observed.violations..Daily.)

gorge_2021_raw <- gorge_2021_raw |>
  mutate(observed_gorge_users = Number.of.observed.gorge.users,
         observed_violations = Number.of.observed.violations)

```

## Data Cleaning

```{r}

clean_gorge <- function(df) {
  df <- df |>
  mutate(Date = as.Date(Date, format('%m/%d/%Y'))) |>
  drop_na(Date) |>
  mutate(High.Temperature = as.numeric(High.Temperature)) |> 
  clean_names()
  
  colnames(df) <- gsub("number\\_of\\_", "", colnames(df))
  
  return(df)
}

# lapply(gorges, function)

gorge_2022_clean <- gorge_2022_raw |>
  clean_gorge() 

gorge_2022_clean <- gorge_2022_clean |>
  mutate(date = replace(date, row_number() == 23, '2022-06-09')) |>
  mutate(steward_name = if_else(steward_name == "Philip", 
                                "Phillip", 
                                steward_name)) |>
  mutate(steward_name = str_replace_all(
    str_to_title(steward_name), 
    " ", 
    "")) 
    

gorge_2022 <- gorge_2022_clean
```
We first import the sets and clean up the data best we can.

# Gorge Statistics
We are interested in understanding the Gorges by the numbers. `Gorges` is a dataset that has all the years joined into one dataset. Due to format of the data changing, some granular detail is lost.

## Join the Gorges

```{r}

gorges_raw <- read.csv('data/gorges.csv')

gorges_clean <- gorges_raw |>
  drop_na(Date) |>
  drop_na(Number.of.observed.gorge.users) |>
  drop_na(total.Number.of.people.obseved.violating.a.rule) |>
  mutate(Date = as.Date(Date, format('%m/%d/%Y')),
         observed_gorge_users = as.numeric(Number.of.observed.gorge.users),
         observed_violations = as.numeric(
           total.Number.of.people.obseved.violating.a.rule),
         High.Temperature = as.numeric(High.Temperature)) |>
  clean_names()

gorges <- gorges_clean |>
  mutate(observed_gorge_users = ifelse(is.na(observed_gorge_users), 
                                       0,
                                       observed_gorge_users),
         observed_violations = ifelse(is.na(observed_violations),
                                      0,
                                      observed_violations),
         high_temperature = ifelse(is.na(high_temperature), 
                                   0,
                                   high_temperature)
         ) |>
  filter(year(date) > (2014))
  
```

## Yearly Stats

```{r}

gorges_tbl <- gorges |>
  group_by(year(date)) |>
  summarise(shifts_completed = n(),
            visitors = sum(observed_gorge_users),
            violations = sum(observed_violations),
            stewards = length(unique(steward_name)),
            avg_temp = mean(high_temperature[high_temperature != 0])) |>
  mutate(year = `year(date)`)

gorges_tbl
```
We see some interesting points. Overall, it seems that the around 14,000 visitors experience the Gorges every year, and around 12 stewards patrol every Summer. The average temperature is a high 70s, and around 100 shifts are completed every year. Notably are the years 2018, 2019, 2020, and 2021.

```{r}

gorges_tbl |> 
  ggplot(aes(x = year, y = visitors)) +
  geom_line()
  
gorges_tbl |> 
  ggplot(aes(x = year, y = violations)) +
  geom_line() + 
  geom_smooth(method = lm, se = F)

gorges_tbl |> 
  ggplot(aes(x = year)) +
  geom_line(aes(y = visitors), color = "blue", linetype = "solid") +
  geom_line(aes(y = violations), color = "red", linetype = "dashed") 


```
Interestingly, we see the amount of visitors spikes in 2016, and falls to it lowest in 2021. 2020 shows only slightly lesser numbers than previous years, and 2022 shows membership returned near fully. It's hard to determine why this drop occured - the COVID-19 Pandemic would be a likely explanation, but that doesn't seem to account for 2020's regular number. 

Over the 8 years of the program, violations overall have a downward trend, which is a good outcome of the Stewards Program. However, the decline is not steady, and there's a noticeable spike in 2020 and 2022. Again, this could be explained by the COVID pandemic: one of the few activities during the pandemic was going outside, and perhaps individuals did not follow rules as they were unaware of them or more careless during that year.


# A Look at 2022

We now choose to look at a specific year. We select 2022 as it has high visitor numbers and is more recent.

```{r}
#| label: point-date-high-temp

gorge_2022 |>
  ggplot(aes(x = date, y = high_temperature)) +
  geom_point() + ylim(0, 100) + geom_line() +
  labs(
    title = 'Temperature Highs over Summer 2022'
  )

gorge_2022 |>
  filter(steward_name != '') |>
  ggplot(aes(x = steward_name)) + 
  geom_histogram(stat = 'Count') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    title = 'Frequency of Shifts by Gorge Steward'
  )

gorge_2022 |>
  ggplot(aes(x = date, y = observed_gorge_users)) +
  geom_point() + geom_line() + ylim(0, 350) + 
  labs(
    title = 'Gorge Visitors over Summer 2022'
  )

gorge_2022 |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = observed_gorge_users), color = "blue", linetype = "solid") +
  geom_line(aes(y = observed_violations), color = "red", linetype = "dashed") +
  labs(title = "Gorge Users and Violations in Summer 2022", 
       x = "Date", 
       y = "Count")
  
gorge_2022 |>
  drop_na(observed_gorge_users) |>
  mutate(dayofweek = weekdays(date)) |>
  group_by(dayofweek) |>
  summarize(weekday_count = sum(observed_gorge_users)) |>
  ggplot(aes(dayofweek, weekday_count)) +
  geom_col() +
  labs(
    title = "Total Visitors by Weekday"
  )

```
# Word Usage
```{r}
library(stringr)

common_prepositions <- c("this", "is", "a", "an", "the", "and", "with", "than", "of", "in", "on", "at", "for", "to", "from", "by", "as", "but", "or", "", "-", "\"hi\"", "\"how", "\"the", "&", "(seemed", "+", "1", "10", "12-1")

most_freq_words <- function(df, column, top_n, ignore_words) {
  df %>%
    pull(!!column) %>%
    str_split("\\s+") %>%
    unlist() %>%
    tolower() %>%
    str_replace_all("[[:punct:]]$", "") %>%  # Remove trailing punctuation marks
    str_remove("\\d+") %>%        
    setdiff(common_prepositions) %>%
    table() %>%
    as.data.frame() %>%
    arrange(desc(Freq)) 
}



interact_words <- most_freq_words(gorge_2022, 'description_of_interactions', 50)


```



Questions to answer:

Does temp affect visitors?
What day of the week or am/pm is busiest?
How have violations gone down?
What words are most frequent?
How did covid affect gorge usage/how has it rebounded?

Future:
Would also like to introduce photos

Public facing installation to highlight safety?
Where is are the most violations occuring?

Safety might be a great approach
Safety report like from perfect match

What do the national parks do that might be similar?

Photographing the beauty of the gorges








